---
title: "Preprocessing"
author: "Daniel Shu"
date: "`r Sys.Date()`"
output: 
  html_document:
    keep_md: yes
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message =TRUE,
                      warning =TRUE, cache= FALSE)
```

# I. Setup
## a. Load libraries

```{r libraries}
library(Seurat)
library(SeuratObject)
library(sctransform)
library(patchwork)
library(Azimuth)
library(SeuratData)
library(ggplot2)
library(tidyverse)
library(glmGamPoi)
library(tools)
library(SeuratDisk)
```

## b.Preliminaries

```{r settings}
data.dir <- "./data" 

list <- list.files(path=data.dir, pattern="tsv.gz|mtx.gz",recursive=TRUE,full.names =TRUE) 
list <- list %>% str_replace_all(c("barcodes.tsv.gz" = "","features.tsv.gz"="","matrix.mtx.gz"="")) %>% unique 

list.names <- list %>% str_replace("./data/","") %>% str_replace("/","")

```

# II. Load data 
### a. Load data
```{r load_seurat}
seurat_list <- lapply(1:length(list),function(i) { 
  tenx_data <- Read10X(data.dir = list[i])
  temp <- CreateSeuratObject(counts = tenx_data, min.cells = 3, min.features = 200) #this creates the seurat objewct
  temp$orig.ident=list.names[i] #this changes the orig.ident column so it contains the name of the sample
  return(temp)
})

seurat <- merge(seurat_list[[1]], y=c(seurat_list[-1]), add.cell.ids=list.names,project="GENEOS")

#add column for patient ID, which is identical to orig.ident except for the two samples where there is an A and B suffix
seurat$Patient <- str_replace(seurat@meta.data$orig.ident, "A|B", "")

# seurat <- readRDS("./output/seurat_before_preprocessing.rds")
# saveRDS(seurat, file = "output/seurat_before_preprocessing.rds")
```

# III. Pre-processing
### a. Plots for percent mito, nFeature
```{r preprocessing}
seurat <-PercentageFeatureSet(seurat, pattern = "^MT-", col.name = "percent.mt")

# Use FeatureScatter is typically used to visualize feature-feature relationships, but can be used
# for anything calculated by the object, i.e. columns in object metadata, PC scores etc.
plot1 <- FeatureScatter(seurat, feature1 = "nCount_RNA", feature2 = "percent.mt", group.by='orig.ident')
plot2 <- FeatureScatter(seurat, feature1 = "nCount_RNA", feature2 = "nFeature_RNA",group.by='orig.ident')
plot1 + plot2

# plot3 <- FeatureScatter(seurat, feature1 = "nCount_RNA", feature2 = "percent.ribo", group.by='orig.ident')
# plot3

# Visualize QC metrics as a violin plot
vln1 <- VlnPlot(seurat, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"#,"percent.ribo"
                                     ),ncol=3)
vln1
vln2 <- VlnPlot(seurat, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"#,"percent.ribo"
                                     ),group.by = 'orig.ident',ncol=3)
vln2

seurat.subset <- subset(seurat, subset = nFeature_RNA > 200 & nFeature_RNA < 4000 & percent.mt < 25) 

vln3 <- VlnPlot(seurat.subset, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
vln4 <- VlnPlot(seurat.subset, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),group.by = 'orig.ident') 
vln3
vln4

pdf("output/qc_plots.pdf");plot1 + plot2;vln1;vln2;vln3;vln4;dev.off()
```

### b. Remove TRA/TRB genes to avoid clonotype bias
Sundell et al. 2023 https://academic.oup.com/bfg/article/22/3/263/6874511

```{r preprocessing}
#subset to remove TRAV and TRBV genes as markers
rownames(seurat.subset) %>% length
grepl("^TR[AB][VJC]", rownames(seurat.subset)) %>% table
seurat.subset <- seurat.subset[!grepl("^TR[AB][VJC]", rownames(seurat.subset)), ] #note I remove TRA TRB genes but not TRG or TRD
rownames(seurat.subset) %>% length

# seurat.subset <- readRDS("./output/seurat_subset.rds")
saveRDS(seurat.subset, file = "output/seurat_subset.rds")

# rm(seurat);rm(seurat_list);rm(plot1);rm(plot2);rm(list=ls(pattern="^vln"))
```

# IV. Normalize Data 
## a. First Attempt wo batch correction
### 1. Run SCTransform. 
NOTE: I am not scaling the data based on this documentation, which says sctransform normalization replaces this function
https://satijalab.org/seurat/articles/sctransform_vignette.html

```{r normalize}
seurat.subset.scT <- SCTransform(seurat.subset,
                                        method="glmGamPoi",
                                        vst.flavor="v2",
                                        vars.to.regress = "percent.mt",
                                 # return.only.var.genes = FALSE #this is set to false so scale.data matrix contains all genes, not just variable genes
                                        )

# save object that has been subset to and scTransformed. If needing to revise can reload this object rather than re-running the above
#saveRDS(seurat.subset.scT, file = "output/seurat_subset_scT.rds")
# seurat.subset.scT <- readRDS("./output/seurat_subset_scT.rds") 
```

### 2. Dimensionality reduction 
This is an amalgamation of the seurat vignettes here (https://satijalab.org/seurat/articles/sctransform_vignette.html) and here (https://satijalab.org/seurat/articles/pbmc3k_tutorial.html)

```{r dimensionality_reduction_find_clusters}
seurat.subset.scT.dimReduction <- RunPCA(seurat.subset.scT)
rm(seurat.subset.scT)

VizDimLoadings(seurat.subset.scT.dimReduction, dims = 1:2, reduction = "pca")
DimPlot(seurat.subset.scT.dimReduction, reduction = "pca")
DimHeatmap(seurat.subset.scT.dimReduction, dims = 1, cells = 500, balanced = TRUE)
DimHeatmap(seurat.subset.scT.dimReduction, dims = 1:15, cells = 500, balanced = TRUE)
ElbowPlot(seurat.subset.scT.dimReduction)

pdf("output/dimreduction_plots.pdf")
VizDimLoadings(seurat.subset.scT.dimReduction, dims = 1:2, reduction = "pca")
DimPlot(seurat.subset.scT.dimReduction, reduction = "pca")
DimHeatmap(seurat.subset.scT.dimReduction, dims = 1, cells = 500, balanced = TRUE)
DimHeatmap(seurat.subset.scT.dimReduction, dims = 1:15, cells = 500, balanced = TRUE)
ElbowPlot(seurat.subset.scT.dimReduction)
dev.off()

seurat.subset.scT.dimReduction <- FindNeighbors(seurat.subset.scT.dimReduction, reduction="pca", dims = 1:15)

#saveRDS(seurat.subset.scT.dimReduction, file = "output/seurat_pre_FindClusters.rds")
# seurat.subset.scT.dimReduction <- readRDS("output/seurat_pre_FindClusters.rds")
```

### 3. Find clusters

```{r find_clusters}
seurat.final <- FindClusters(object = seurat.subset.scT.dimReduction, resolution = 0.7)
head(Idents(seurat.final), 5)
seurat.final <- RunUMAP(seurat.final, reduction="pca", dims=1:15) 

#Visualize with UMAP
p1<- DimPlot(seurat.final, reduction = "umap")
p1
p2 <- DimPlot(seurat.final, repel =TRUE, split.by="orig.ident", ncol=3)
p2
p3 <- DimPlot(seurat.final, repel =TRUE, group.by="orig.ident")
p3
p2/p3
pdf("output/find_clusters.pdf");p1;p2;p3;p2/p3;dev.off();rm(p1);rm(p2);rm(p3)

saveRDS(seurat.final, file = "output/seurat_final.rds")
# seurat.final <- readRDS("./output/seurat_final.rds")
# rm(list=ls(pattern="^seurat.subset"))
```

# V. Normalize Data + batch correction

In section IV.1, I determined that there was significant batch effect between sample 8A and 8B vs 6 and 7, so here I am adding batch correction

```{r}
# seurat.subset <- readRDS("./output/seurat_subset.rds")

# now will integrate following this tutorial (https://satijalab.org/seurat/articles/integration_introduction.html)
# note this comes from the bottom of the page, where they demonstrate integration for datasets using scTransform normalization
# so i will go back to the output of section III (after QC to remove cells with high mitochondrial content) and split that object, seurat.subset, which is analogous to "ifnb" in the seurat vignette


#  I went to a reference-based integration by orig.ident (see FindIntegrationAnchors below)
#note I use ifnb object names since this code is copied from the vignette
obj.list <- SplitObject(seurat.subset, split.by = "orig.ident")
obj.list <- lapply(X = obj.list, FUN = SCTransform, method = "glmGamPoi",vst.flavor="v2",vars.to.regress = "percent.mt")
features <- SelectIntegrationFeatures(object.list = obj.list, nfeatures = 3000)
obj.list <- PrepSCTIntegration(object.list = obj.list, anchor.features = features)
obj.list <- lapply(X = obj.list, FUN = RunPCA, features = features)

########### perform reference based integration
#(https://satijalab.org/seurat/articles/integration_large_datasets.html)
#other type of integration causes memory error
anchors <- FindIntegrationAnchors(object.list = obj.list, reference = 1, 
                                  normalization.method = "SCT", anchor.features=features,
                                  reduction = "rpca", dims = 1:30)
saveRDS(anchors, file = "./output/anchors.rds")

immune.combined.sct <- IntegrateData(anchorset = anchors, normalization.method = "SCT", dims = 1:30) #this step is ~3 minutes

#workflow after integration based on this: https://github.com/satijalab/seurat/issues/1836
#########
immune.combined.sct <- RunPCA(immune.combined.sct, verbose = FALSE)
immune.combined.sct <- RunUMAP(immune.combined.sct, reduction = "pca", dims = 1:15)

pdf("output/dim_attempts.pdf")
for (i in 10:30) {
  immune.combined.sct <- RunUMAP(immune.combined.sct, reduction = "pca", dims = 1:i)
  px <- DimPlot(immune.combined.sct, reduction = "umap") +ggtitle(paste0(i, " Dims"))
  py <- DimPlot(immune.combined.sct, reduction = "umap",split.by="orig.ident")
  print(px/py&NoLegend())
  
  immune.combined.sct <- FindNeighbors(immune.combined.sct,reduction="pca",dims=1:i) 
  #this isrequired to run findclusters on the combined object. see https://github.com/satijalab/seurat/issues/2995
  seurat.combined.final <- FindClusters(object = immune.combined.sct, resolution = 0.7)
  p1<- DimPlot(seurat.combined.final, reduction = "umap",label=TRUE)+ggtitle(paste0(i, " Dims"))
  p3 <- DimPlot(seurat.combined.final, repel =TRUE, split.by="orig.ident")
  print(p1/p3&NoLegend())
}
dev.off()

#after reviewing results of above, decided to go with 10 dims
pdf("output/dim_final.pdf")
for (i in 10) {
  immune.combined.sct <- RunUMAP(immune.combined.sct, reduction = "pca", dims = 1:i)
  px <- DimPlot(immune.combined.sct, reduction = "umap") +ggtitle(paste0(i, " Dims"))
  py <- DimPlot(immune.combined.sct, reduction = "umap",split.by="orig.ident")
  p0 <- DimPlot(immune.combined.sct,reduction="pca")
  print(p0)
  print(px/py&NoLegend())
  
  immune.combined.sct <- FindNeighbors(immune.combined.sct,reduction="pca",dims=1:i) 
  #this isrequired to run findclusters on the combined object. see https://github.com/satijalab/seurat/issues/2995
  seurat.combined.final <- FindClusters(object = immune.combined.sct, resolution = 0.7)
  p1<- DimPlot(seurat.combined.final, reduction = "umap",label=TRUE)+ggtitle(paste0(i, " Dims"))
  p3 <- DimPlot(seurat.combined.final, repel =TRUE, split.by="orig.ident")
  print(p1/p3&NoLegend())
}
dev.off()

# saveRDS(immune.combined.sct, file = "output/immune_combined_sct.rds")
immune.combined.sct <- readRDS("./output/immune_combined_sct.rds")
# saveRDS(seurat.combined.final, file = "output/seurat_combined_final.rds")
seurat.combined.final <- readRDS("./output/seurat_combined_final.rds")

rm(anchors)
rm(obj.list)
rm(list=ls(pattern="p.*"))
rm(seurat.subset)
```

# VI. Multimodal reference mapping

### a. Map to pbmc_multimodal.h5seurat

https://satijalab.org/seurat/articles/multimodal_reference_mapping.html

note this reference supports our integration of the 4 datasets and removal of batch effect before mapping
https://github.com/satijalab/seurat/issues/3686

```{r}
reference <- LoadH5Seurat("data/pbmc_multimodal.h5seurat") # compiling this reference object takes a very long time, so here I load a saved copy
# saveRDS(reference,file="output/pbmc_multimodal_reference_as_rds.rds")
# reference<-readRDS("output/pbmc_multimodal_reference_as_rds.rds") 


anchors <- FindTransferAnchors(
  reference = reference,
  query = seurat.combined.final,
  normalization.method = "SCT",
  reference.reduction = "spca",
  dims = 1:50
)
saveRDS(anchors,file="output/findTransferAnchors_output.rds")

# # the mapquery function below doesn't work with uwot 0.1.14 
# # per this page https://github.com/satijalab/seurat/issues/6345
# # so needed to install older version
# require(devtools)
# # remove.packages("uwot"
# install_version("uwot", version="0.1.11",repos = "http://cran.us.r-project.org")

# then run
seurat.combined.final <- MapQuery(
  anchorset = anchors,
  query = seurat.combined.final,
  reference = reference,
  refdata = list(
    celltype.l1 = "celltype.l1",
    celltype.l2 = "celltype.l2",
    celltype.l3 = "celltype.l3",
    predicted_ADT = "ADT"
  ),
  reference.reduction = "spca", 
  reduction.model = "wnn.umap"
)

# saveRDS(seurat.combined.final,file="output/seurat_combined_final_afterMapQuery.rds")
# seurat.combined.final <- readRDS("output/seurat_combined_final_afterMapQuery.rds")

```

### b. Make plots

```{r}
p1 <- DimPlot(seurat.combined.final, reduction = "ref.umap", group.by = "predicted.celltype.l1", label = TRUE, label.size = 3, repel = TRUE) + NoLegend()
p2 <- DimPlot(seurat.combined.final, reduction = "ref.umap", group.by = "predicted.celltype.l2", label = TRUE, label.size = 3 ,repel = TRUE) + NoLegend()
p3 <- DimPlot(seurat.combined.final, reduction = "ref.umap", group.by = "predicted.celltype.l3", label = TRUE, label.size = 3 ,repel = TRUE) + NoLegend()
p1 + p2  
p3

p3.5 <- DimPlot(seurat.combined.final, reduction = "ref.umap", group.by = "predicted.celltype.l2", split.by="orig.ident",label = FALSE,repel = TRUE) + NoLegend()
p3.5

pdf("output/seurat_combined_final_afterMapQuery_plots.pdf");p1;p2;p3;p3.5;dev.off()

p4<-FeaturePlot(seurat.combined.final, features = c("CD8 TEM", "CD4 CTL", "Treg"),  reduction = "ref.umap", cols = c("lightgrey", "darkred")) & theme(plot.title = element_text(size = 10))
p4

Idents(seurat.combined.final) <- 'predicted.celltype.l3'
levels(seurat.combined.final) <-  levels(seurat.combined.final) %>% sort

DefaultAssay(seurat.combined.final) <- 'SCT'
p5<-VlnPlot(seurat.combined.final, features = c("CD3E", "CD4", "CD8B", #"CD69",
                                                # "CXCR3","cXCR5", 
                                                "SELL", "CCR7", "TCF7", 
                                                "GZMK","GZMB","GNLY", "NKG7",#"ZNF683",
                                                "CTLA4","PDCD1"), 
            idents=Idents(seurat.combined.final)[grep("CD4|CD8|NK|dnT|gdT|MAIT|Treg",Idents(seurat.combined.final))],
            stack=TRUE,sort = FALSE,flip=TRUE) + 
  theme(axis.text.x = element_text(size=8))+NoLegend()
p5

goi <- c(#"IDO1", 
         "NKG7", "PDCD1","LCK", "CCL5", "IFNG", "CXCR6", "CD27", "CD69", "CD8A", "CD8B", "CD3E", "CD3G", "TIGIT")
p5.5<-VlnPlot(seurat.combined.final, features = goi, 
            idents=Idents(seurat.combined.final)[grep("CD4|CD8|NK|dnT|gdT|MAIT|Treg",Idents(seurat.combined.final))],
            stack=TRUE,sort = FALSE,flip=TRUE) + theme(axis.text.x = element_text(size=8))+NoLegend()+ggtitle("GOI from RP 06-20-2023",subtitle="IDO1 excluded due to same value across all cells")
p5.5

DefaultAssay(seurat.combined.final) <- 'predicted_ADT'
# see a list of proteins: rownames(pbmc3k)
p6<-FeaturePlot(seurat.combined.final, features = c("CD3-1", "CD45RA", "IgD"), reduction = "ref.umap", cols = c("lightgrey", "darkgreen"), ncol = 3)
p6

pdf("output/seurat_combined_final_afterMapQuery_plots2.pdf");p4;p5;p5.5;p6;dev.off()

############ alternative viz
DefaultAssay(seurat.combined.final) <- 'SCT'

pdf("output/seurat_combined_final_afterMapQuery_plots_MyUmap.pdf")
lapply(1:3, function(x) {
  DimPlot(seurat.combined.final, reduction = "umap", 
          group.by = paste0("predicted.celltype.l", x), 
          label=TRUE,
          repel = TRUE) + NoLegend()
 
});DimPlot(seurat.combined.final, reduction = "umap", 
          split.by="orig.ident",
          repel = TRUE) + NoLegend()
dev.off()
```

# VII. Session info
```{r sessioninfo}
sessionInfo()
writeLines(capture.output(sessionInfo()), "sessionInfo_1_preprocessing.txt")
```
